---
title: "Modeling price of homes"
author: "Luke Jimenez"
date: "12/06/2023"
bibliography: PackagesUsed.bib
output: 
  bookdown::html_document2: 
    highlight: textmate
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
# Data Set

housedata <- read.csv("https://lasanthi-asu.github.io/STT3851ClassRepo/Rmarkdown/Data/housedata.csv", 
                      colClasses = c(id = "character", date = "character", 
                                     yr_built = "character", zipcode = "factor", grade = "factor"))

housedata$date <- as.Date(housedata$date, "%Y%m%d")
housedata$waterfront <- factor(housedata$waterfront, labels = c("No", "Yes"))
housedata$condition <- factor(housedata$condition, labels = c("poor", "fair", "average", "good", "very good"))
housedata$yr_renovated <- ifelse(housedata$yr_renovated == 0, housedata$yr_built, housedata$yr_renovated)
housedata$yr_built <- as.Date(ISOdate(housedata$yr_built, 9, 1))  # Complete Year, Sept 1
housedata$yr_renovated <- as.Date(ISOdate(housedata$yr_renovated, 9, 1))  # Last renovated Year, Sept 1

housedata <- housedata[, -1]


# Test Data Set (only use this to send predictions)

#### Perform same steps with test set

housedataT <- read.csv("https://lasanthi-asu.github.io/STT3851ClassRepo/Rmarkdown/Data/housedataTEST.csv", 
                      colClasses = c(id = "character", date = "character", 
                                     yr_built = "character", zipcode = "factor", grade = "factor"))

housedataT$date <- as.Date(housedataT$date, "%Y%m%d")
housedataT$waterfront <- factor(housedataT$waterfront, labels = c("No", "Yes"))
housedataT$condition <- factor(housedataT$condition, labels = c("poor", "fair", "average", "good", "very good"))
housedataT$yr_renovated <- ifelse(housedataT$yr_renovated == 0, housedataT$yr_built, housedataT$yr_renovated)
housedataT$yr_built <- as.Date(ISOdate(housedataT$yr_built, 9, 1))  # Complete Year, Sept 1
housedataT$yr_renovated <- as.Date(ISOdate(housedataT$yr_renovated, 9, 1))  # Last renovated Year, Sept 1
housedataT <- housedataT[, -1]
```

# Introduction

The Department of Assessments in King County stores a significant amount of data on the homes within the county. In this paper, we will be using the different characteristics or predictor variables listed within this data in order to estimate the price of the homes. Creating a model to predict price will help the reader understand the relationship of price to the predictor variables, and how significant each predictor is. The analysis of the subset selection reveals a trend of a large amount of significant variables, and the analysis of the model furthermore reveals the importance of categorical variables in predicting price.

# Methods

## Data Collection

The data used in this paper was already curated and downloaded from

<https://lasanthi-asu.github.io/STT3851ClassRepo/Rmarkdown/Data/housedata.csv>

It is not clear when and how the data was collected. The only information given is the origin of the data, King County Department of Assessments.

## Exploratory Analysis

Exploratory analysis was performed by examining scatter plots, residual plots, and subset selection. In order to correct a positive skew in the residual plot, we applied a base 10 logarithm to price. Base 10 logarithms were added as well to certain predictor variables as it helped adjust the model to the new values and increase the adjusted $R^2$ values. These predictor variables include square footage of the apartments interior living space (`sqft_living`), square footage of the land space (`sqft_lot`), and the square footage of interior housing living space for the nearest 15 neighbors (`sqft_living15`).

```{r, echo = FALSE, fig.height = 2, fig.width = 3, fig.align ='center'}
model1 <- lm(price~bedrooms + sqft_living + waterfront + grade + sqft_living15 + sqft_lot15 + condition + zipcode +long, data = housedata)

p1<- ggplot(model1, aes(x = .fitted, y = .resid)) + geom_point(color = "red", size = 0.5) + theme_bw() +xlab("Fitted Values")+ ylab("Residuals")
p1

```

Figure 2.1: The residual plot of the final price model before applying base 10 logarithms to `price`, `sqft_living`, `sqft_lot`, `sqft_living15`. The adjusted $R^2$ value is *0.8228*

```{r, echo = FALSE, fig.height = 2, fig.width = 3, fig.align ='center'}
model2 <- lm(log(price)~  log(sqft_living) + waterfront + grade + log(sqft_living15) + log(sqft_lot15) + condition + zipcode + long + view + bedrooms + bathrooms + lat + date , data = housedata)


p2<- ggplot(model2, aes(x = .fitted, y = .resid)) + geom_point(color = "red", size = 0.5) + theme_bw()+ xlab("Fitted Values")+ ylab("Residuals")
p2

```

Figure 2.2: The residual plot of the final price model after applying base 10 logarithms to `price`, `sqft_living`, `sqft_lot`, `sqft_living15`. The adjusted $R^2$ value is *0.8833*

## Statistical Modeling

Statistical concepts such as forward subset selection, analysis of residual plots, and multivariate regression techniques.

## Reproducibility

All analyses performed in this paper can be reproduced by running the original `.Rmd` file with RStudio, assuming the link to the original data remains current and the contents thereof remain unchanged. The `R` packages `ggplot2` (Wickham et al. 2020) and `knitr` (Xie 2021) will need to be installed on the user's computer.

# Results

The variables used to develop the final price prediction model includes square footage of the apartments interior living space (`sqft_living`), square footage of the land space (`sqft_lot`), and the square footage of interior housing living space for the nearest 15 neighbors (`sqft_living15`), lattitude (`lat`), longitude (`long`), the zipcode area the house is in (`zipcode`), an index from 1 to 13 where 1-3 falls short of building construction and design, 7 has an average level of construction and design, and 11-13 have a high quality level of construction and design (`grade`), a dummy variable determining if the apartment is overlooking the waterfront (`waterfront`), an index from 0 to 4 of how good the view of the property was, 0 - lowest, 4 - highest (`view`), an index from 1 to 5 on the condition of the apartment, 1 - lowest, 4 - highest (`condition`), number of bedrooms (`bedrooms`), and the number of bathrooms (`bathrooms`).

The final model used was

$$\beta_0 + \beta_{1}log10(sqft\_lot) + \beta_{2}log10(sqft\_living) + \beta_{3}log10(sqft\_living15) +  \beta_4(lat) + \beta_5(waterfront) + \beta_6(view) + \\  \beta_7(bedrooms) +  \beta_8(bathrooms)+  \beta_{9}f(grade:3) + \dots \beta_{19} f(grade:13)+  \beta_{20} f(condition:fair) + \dots \beta_{23}f(condition:very\_good)\\ + \beta_{24}(zipcode:98002) + \dots \beta_{92}(zipcode:98199)+ \epsilon$$

The variables `f(zipcode)`, `f(grade)`, and `f(condition)` are categorical predictor variables of varying levels.

```{r, echo = FALSE, fig.height = 2, fig.width = 5, fig.align ='center', warning = FALSE}
suppressMessages({
p3<- ggplot(model2, aes(x = .fitted, y = .resid)) + geom_point(color = "pink", size = 0.5) + theme_bw()+ xlab("Fitted Values")+ ylab("Residuals")
p3
})
```

Figure 3.1: The graph shows the residuals of the final models predictions of the training data points. Note that there is no pattern in the residual plot, indicating that the multivariate regression model created is an appropriate one.

There is a highly statistically significant relationship ($p$-value \$ \< 0.0001\$) between the base 10 logarithm of price and all the variables in the model, excluding certain levels of the three categorical variables

# Conclusions

The goal of the analysis was to minimize the test $MSPE^{1/2}$. Without access to the test data, the criterion we used to decide on a model was the adjusted $R^2$ value. Throughout the creation of the model we used best subset selection, forward selection, and lasso shrinkage in order to create a good model. In order to be computationally efficient, the forward selection model worked best for subset selection. Through analysis of residual plots of single variable linear models and correlation, we applied base 10 logarithms to `price`, `sqft_living`, `sqft_lot`, `sqft_living15`. The final test $MSPE^{1/2}$ was 123730.8

```{r, comment = NA, results = "hide", messages = FALSE, echo = FALSE}
suppressMessages({PackagesUsed <- c( "ggplot2", "ISLR", "knitr", "dplyr", "rmarkdown", "bookdown")
# Write bib information
knitr::write_bib(PackagesUsed, file = "./PackagesUsed.bib")
# Load packages
lapply(PackagesUsed, library, character.only = TRUE)})
```
## References


This document uses `ggplot2` by @R-ggplot2, `ISLR` by @R-ISLR, `rmarkdown` by @R-rmarkdown, `dplyr` by @R-dplyr, `knitr` by @R-knitr, and `bookdown` by @R-bookdown.
